Supplementary materials for 'The collapse of eastern Bering Sea snow crab

### 2023-07-21 manuscript_supmat.Rmd
third revision

#-  \pagenumbering{gobble}
csl: fish-and-fisheries.csl
toc: yes
title: "Supplementary materials for 'Canaries of the Arctic: the collapse of eastern Bering Sea snow crab'"
---

@@

## Population dynamics model
The population dynamics model presented here incorporated the best available information on relevant population processes to estimate total mortality for male snow crab on the eastern Bering Sea shelf and is similar in structure to the model used to assess eastern Bering Sea snow crab for management (Szuwalski, 2021). The model tracked numbers of male crab at size at maturity state over time with size bins ranging from 30-95 mm carapace width with 5 mm bin widths. Only male crab were modeled because male and female crab appear to have somewhat different dynamics and the male crab in the modeled size range are better selected by the survey gear (Szuwalski, 2021). Snow crab are sexually dimorphic, with male snow crab growing to nearly twice the size of females, which accounts for the better selection in the survey. Only crab smaller than 95 mm were modeled for two reasons: 1) to attempt to isolate the effect of the directed fishery (crabs of >101 mm carapace width are targeted in the fishery; discussed further below) and 2) almost all of the crab that disappeared since 2018 are in this size range. The population dynamics model operates on a half year time step, starting in July at the time of the NMFS survey.  Total mortality (Z) is estimated by year (y) and maturity state (m). Other estimated parameters include the initial numbers at size by maturity state, yearly log recruitments, a vector of scalars that determine the proportions of estimated recruitment split into the first two size bins, and a variance component for the penalty on total mortality. Parameters determining growth, maturity, and survey selectivity were estimated outside of the model and specified when estimating mortality and catchability. Mortality is the only population process that occurs in the first half of a given year:

@@

Cannibalism has been proposed as a potential driver of the dynamics of snow crab in eastern Canada (Lovrich et al., 1997). In laboratory studies, crab smaller than 55 mm carapace width were at high risk of being cannibalized when housed with larger crab (Lovrich et al., 1997). Crab larger than 55 mm carapace width were much less likely to be cannibalized, but the frequency of injury could be high. Here we developed an index of cannibalism based on two aspects of the spatial distribution of snow crab: the overlap of crab smaller than 55 mm carapace width with crab larger than 95 mm carapace width (\autoref{distribution_males_le55}) and the density of crab larger than 95 mm carapace width within the shared space. The proportion of 55 mm carapace width crab in the overlapping area represents the 'exposure' of the smaller population to cannibalism and the density of crab larger than 95 mm carapace width within that area represents the potential 'intensity' of cannibalism in the shared area. We calculated an index of cannibalism over time as the product of exposure and intensity. Consequently, a scenario in which there was large overlap, but low densities of large crab would result in a low cannibalism index value. Similarly, a scenario in which there was low overlap, but high densities would result in a low cannibalism index value. This produces an index that is comparable with estimated mortality--a higher cannibalism index would be expected to be associated with higher mortality if cannibalism is a strong driver of mortality in the size ranges of crabs modeled here.

@@

## Sensitivities to model assumptions in GAMs
Modeling decisions are also necessarily made in the process of fitting GAMs and it is possible for these decisions to influence the outcome of an analysis. Within the context of our model, these decisions include the assumed error structure, the treatment of the uncertainty associated with the estimates of mortality, and the allowed shape of the smooths estimated in the GAMs. The following sensitivities address the implications of these modeling decisions on the outcome of our analysis.

### Error structure
Impacts of assumptions about error structure were explored by assuming beta distributed data in the GAM and transforming the continuous total mortality rates to an exploitation rate ranging from 0 to 1. This transformation still resulted in mature population and temperature being the most important variables related to mortality, however the deviance explained decreased to 60% and 68% for mature and immature mortality, respectively, compared to 72% and 78% for the model presented in the main text. A potential shortcoming of the method presented in the main text is that predicted mortality could be less than zero. This was not the case in any of the model fittings, and would present a problem primarily if the model was extrapolated to data beyond the observed ranges. A potential fix to this issue is to log the response variable, but this resulted in unacceptable patterns in the residuals, so this model was not used.

@@

## A word on methods
Attribution of changes in population processes in ecology is a difficult problem, particularly for wild populations that are difficult to directly observe and impossible to experiment on in situ. There are a large range of methodologies that claim to identify causality in observational data (structural equation modeling, empirical dynamic modeling, etc.). Some of the difficulties in determining causality in ecological time series are related to the generally short time series that are available, non-linear dynamics, and departures of populations or covariates into unexplored parameter space. These issues can present issues for any modeling framework and we have tried to address these the the best of our ability with the models used here. The use of p-values has been (rightfully) criticized in the literature and the explanatory power of our models are likely overstated given these criticisms.  Ultimately, the numerous sensitivities and simulation tests performed here were undertaken to try to understand if a suite of covariates appear to be important under different modeling decisions and considerations of uncertainty, in spite of the potential short-comings of the data available and models selected. Temperature and population density appear to be these covariates.

@@

### Are you sure the collapse wasn't a result of trawling?
The bycatch index steadily declined since the beginning of our study, with the relative impact of trawling in 2018 and 2019 below the historical average (\autoref{grid_covar}). It is difficult to reconcile the idea that trawling could have contributed to the collapse with the relatively low mortality rates estimated during the periods when the bycatch index was many times higher in the 1990s. Furthermore, if trawling were a large source of mortality for snow crab, it is difficult to understand how the largest pseudocohort ever observed could have established and survived for ~8 years on the Bering Sea shelf, during which the trawling pressure was relatively consistent.

However, not all of the mortality associated with non-directed fleets is observed. The index used in this analysis is a reliable indicator of the trend in bycatch mortality provided the ratio of observed to non-observed mortality is consistent over time. If this is not the case, that could change the outcome of our analysis, but there is no clear methodology for determining that ratio.

@@

```{r,echo=FALSE,warning=FALSE,message=F,out.width="98%",fig.cap="\\label{fish_cpue}Fishery cpue (top; black lines are median, grey box represents 25-75th quantiles, circles are outliers) and number of crab caught in the directed snow crab fishery (bottom). Vertical dashed line represents the introduction of individual transferrable quota management. CPUEs before and after are difficult to compare as a result of changes in fleet behavior."}

@@

```{r,echo=FALSE,warning=FALSE,message=F,out.width="98%",fig.cap="\\label{split_mat}Size at which half of the crab in the population are mature over time (note, this is not the probability of undergoing terminal molt, rather the proportion of the number of mature vs. immature crab at size in the population)."}

@@

```{r,echo=FALSE,warning=FALSE,message=F,out.width="98%",fig.cap="\\label{temp_occupied} Temperature occupied over time of crab by 5 mm size bin. "}

@@

```{r,echo=FALSE,warning=FALSE,message=F,out.width="98%",fig.cap="\\label{temp_by_mat}Temperature occupied over time of crab by maturity state defined by the size at which 50\\% of crab are mature according to chela height. "}

>>>>

#-  \pagenumbering{gobble}
csl: fish-and-fisheries.csl
toc: yes
title: "Supplementary materials for 'The collapse of eastern Bering Sea snow crab'"
---

@@

## Population dynamics model
The population dynamics model presented here incorporated the best available information on relevant population processes to estimate total mortality for male snow crab on the eastern Bering Sea shelf and is similar in structure to the model used to assess eastern Bering Sea snow crab for management (Szuwalski, 2021). The model tracked numbers of male crab at size at maturity state over time with size bins ranging from 30-95 mm carapace width with 5 mm bin widths. Only male crab were modeled because male and female crab appear to have somewhat different dynamics and the male crab in the modeled size range are better selected by the survey gear (Szuwalski, 2021). Snow crab are sexually dimorphic, with male snow crab growing to nearly twice the size of females, which accounts for the better selection in the survey. Only crab smaller than 95 mm were modeled for two reasons: 1) to attempt to isolate the effect of the directed fishery (crab of >101 mm carapace width are targeted in the fishery; discussed further below) and 2) almost all of the crab that disappeared since 2018 are in this size range. The population dynamics model operates on a half year time step, starting in July at the time of the NMFS survey.  Total mortality (Z) is estimated by year (y) and maturity state (m). Other estimated parameters include the initial numbers at size by maturity state, yearly log recruitments, a vector of scalars that determine the proportions of estimated recruitment split into the first two size bins, and a variance component for the penalty on total mortality. Parameters determining growth, maturity, and survey selectivity were estimated outside of the model and specified when estimating mortality and catchability. Mortality is the only population process that occurs in the first half of a given year:

@@

Cannibalism has been proposed as a potential driver of the dynamics of snow crab in eastern Canada (Lovrich et al., 1997). In laboratory studies, crab smaller than 55 mm carapace width were at high risk of being cannibalized when housed with larger crab (Lovrich et al., 1997). Crab larger than 55 mm carapace width were much less likely to be cannibalized, but the frequency of injury could be high. Here we developed an index of cannibalism based on two aspects of the spatial distribution of snow crab: the overlap of crab smaller than 55 mm carapace width with crab larger than 95 mm carapace width (\autoref{distribution_males_le55}) and the density of crab larger than 95 mm carapace width within the shared space. The proportion of 55 mm carapace width crab in the overlapping area represents the 'exposure' of the smaller population to cannibalism and the density of crab larger than 95 mm carapace width within that area represents the potential 'intensity' of cannibalism in the shared area. We calculated an index of cannibalism over time as the product of exposure and intensity. Consequently, a scenario in which there was large overlap, but low densities of large crab would result in a low cannibalism index value. Similarly, a scenario in which there was low overlap, but high densities would result in a low cannibalism index value. This produces an index that is comparable with estimated mortality--a higher cannibalism index would be expected to be associated with higher mortality if cannibalism is a strong driver of mortality in the size ranges of crab modeled here.

@@

## Sensitivities to model assumptions in GAMs
Modeling decisions are also necessarily made in the process of fitting GAMs and it is possible for these decisions to influence the outcome of an analysis. Within the context of our model, these decisions include the assumed error structure, the treatment of the uncertainty associated with the estimates of mortality, and the allowed shape of the smooths estimated in the GAMs. The following sensitivities address the implications of these modeling decisions on the outcome of our analysis.  
### Error structure
Impacts of assumptions about error structure were explored by assuming beta distributed data in the GAM and transforming the continuous total mortality rates to an exploitation rate ranging from 0 to 1. This transformation still resulted in mature population and temperature being the most important variables related to mortality, however the deviance explained decreased to 60% and 68% for mature and immature mortality, respectively, compared to 72% and 78% for the model presented in the main text. A potential shortcoming of the method presented in the main text is that predicted mortality could be less than zero. This was not the case in any of the model fittings, and would present a problem primarily if the model was extrapolated to data beyond the observed ranges. A potential fix to this issue is to log the response variable, but this resulted in unacceptable patterns in the residuals, so this model was not used.

@@

## A word on methods
Attribution of changes in population processes in ecology is a difficult problem, particularly for wild populations that are difficult to directly observe and impossible to experiment on in situ. There are a large range of methodologies that claim to identify causality in observational data (structural equation modeling, empirical dynamic modeling, etc.). Some of the difficulties in determining causality in ecological time series are related to the generally short time series that are available, non-linear dynamics, and departures of populations or covariates into unexplored parameter space. These issues can present problems for any modeling framework and we have tried to address these to the best of our ability with the models used here. The use of p-values has been (rightfully) criticized in the literature and the explanatory power of our models are likely overstated given these criticisms. Ultimately, the numerous sensitivities and simulation tests performed here were undertaken to try to understand if a suite of covariates appear to be important under different modeling decisions and considerations of uncertainty, in spite of the potential short-comings of the data available and models selected. Temperature and population density appear to be these covariates.

@@

### Are you sure the collapse wasn't a result of trawling?
The bycatch index steadily declined since the beginning of our study, with the relative impact of trawling in 2018 and 2019 below the historical average (\autoref{grid_covar}). It is difficult to reconcile the idea that trawling could have contributed to the collapse with the relatively low mortality rates estimated during the periods when the bycatch index was many times higher in the 1990s. Furthermore, if trawling were a large source of mortality for snow crab, it is difficult to understand how the largest pseudocohort ever observed could have established and survived for ~8 years on the Bering Sea shelf, during which the trawling pressure was relatively consistent. It should be noted that trawling occurred farther north in 2018 than usual and some of this effort overlapped with the area in which the largest densities of crab were observed and lost (\autoref{bycatch_binned}). However, the intensity of trawling in that area and the observed bycatch of snow crab in the areas of highest density were very small compared to other areas in the Bering Sea and historical values (\autoref{bycatch_binned_2018}).

Not all of the mortality associated with non-directed fleets is observed. It is possible that crab on the seafloor are impacted by the trawling gear and die as a result of their injuries. The index used in this analysis is a reliable indicator of the trend in bycatch mortality provided the ratio of observed to non-observed mortality is consistent over time. If this is not the case, that could change the outcome of our analysis, but there is no clear methodology for accurately determining that ratio.

@@

```{r,echo=FALSE,warning=FALSE,message=F,out.width="98%",fig.cap="\\label{fish_cpue}Fishery CPUE (top; black lines are median, grey box represents 25-75th quantiles, circles are outliers) and number of crab caught in the directed snow crab fishery (bottom). Vertical dashed line represents the introduction of individual transferrable quota management. CPUEs before and after are difficult to compare as a result of changes in fleet behavior."}

@@

```{r,echo=FALSE,warning=FALSE,message=F,out.width="98%",fig.cap="\\label{split_mat}Size (in mm carapace width) at which half of the crab in the population are mature over time (note, this is not the probability of undergoing terminal molt, rather the proportion of the number of mature vs. immature crab at size in the population)."}

@@

```{r,echo=FALSE,warning=FALSE,message=F,out.width="98%",fig.cap="\\label{temp_occupied} Average temperature occupied over time of crab by 5 mm size bin. "}

@@

```{r,echo=FALSE,warning=FALSE,message=F,out.width="98%",fig.cap="\\label{temp_by_mat}Average temperature occupied over time of crab by maturity state defined by the size at which 50\\% of crab are mature according to chela height. "}
